Appendix

### 2025-03-21 manuscript_rmarkdown/manuscript_appendix.Rmd
Update p-values for empirical tests.

data <- .plot_experimental_invariance(plot_theme = plot_theme, method = "box_cox")
data$p
```

@@

Applying conventional and invariant Box-Cox transformations to age of patients with lung cancer [@Loprinzi1994-cd] yielded the following results:
no transformation (sum of residuals with normal distribution $\sum r_i = 16.5$);
conventional transformation ($\lambda = 1.9$, $\sum r_i = 11.5$, $\mu_{BC} = 1.6 \cdot 10^3$, $\sigma_{BC} = 0.4 \cdot 10^3$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 1.9$, $\sum r_i = 11.5$, $\mu_{BC} = 1.6 \cdot 10^3$, $\sigma_{BC} = 0.4 \cdot 10^3$);
location- and scale-invariant transformation ($\lambda = 1.7$, $\sum r_i = 11.6$, $\mu_{BC} = 1.9$, $\sigma_{BC} = 0.8$);
and robust location- and scale-invariant transformation ($\lambda = 1.5$, $\sum r_i = 11.6$, $\mu_{BC} = 3.6$, $\sigma_{BC} = 1.2$).

@@

Applying conventional and invariant Box-Cox transformations to the body mass of penguins [@Gorman2014-eo] yielded the following results:
no transformation (residual sum $\sum r_i = 48.0$);
conventional transformation ($\lambda = -0.5$, $\sum r_i = 32.2$, $\mu_{BC} = 2.1$, $\sigma_{BC} = 4 \cdot 10^{-3}$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = -0.5$, $\sum r_i = 32.2$, $\mu_{BC} = 2.1$, $\sigma_{BC} = 4 \cdot 10^{-3}$);
location- and scale-invariant transformation ($\lambda = 0.5$, $\sum r_i = 27.3$, $\mu_{BC} = 0.3$, $\sigma_{BC} = 0.6$);
and robust location- and scale-invariant transformation ($\lambda = 0.2$, $\sum r_i = 25.2$, $\mu_{BC} = 0.4$, $\sigma_{BC} = 0.6$).

@@

Applying conventional and invariant Box-Cox transformations to the latitude of houses in the Ames housing dataset [@De-Cock2011-jf] yielded the following results:
no transformation (residual sum $\sum r_i = 328$);
conventional transformation ($\lambda = 62.1$, $\sum r_i = 319$, $\mu_{BC} = 1.1 \cdot 10^{99}$, $\sigma_{BC} = 0.0 \cdot 10^{99}$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 96.0$, $\sum r_i = 319$, $\mu_{BC} = 6.2 \cdot 10^{153}$, $\sigma_{BC} = 0.3 \cdot 10^{153}$);
location- and scale-invariant transformation ($\lambda = 1.9$, $\sum r_i = 312$, $\mu_{BC} = 2.3$, $\sigma_{BC} = 0.9$);
and robust location- and scale-invariant transformation ($\lambda = 1.2$, $\sum r_i = 316$, $\mu_{BC} = 5.5$, $\sigma_{BC} = 1.4$).

@@

data <- .plot_experimental_outlier_robustness(plot_theme = plot_theme, method = "box_cox")
data$p
```

@@

The Top Gear dataset contains data on 297 vehicles, with outliers related highly fuel-efficient vehicles [@Alfons2021-kc].
Applying conventional and invariant Box-Cox transformations to the fuel consumption feature yielded the following results:
no transformation (residual sum $\sum r_i = 54$, $p=0.76$);
conventional transformation ($\lambda = -0.1$, $\sum r_i = 55$, $\mu_{BC} = 3.0$, $\sigma_{BC} = 0.3$, $p=0.01$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 0.8$, $\sum r_i = 48$, $\mu_{BC} = 29$, $\sigma_{BC} = 15$, $p=0.55$);
location- and scale-invariant transformation ($\lambda = -0.7$, $\sum r_i = 44$, $\mu_{BC} = 0.6$, $\sigma_{BC} = 0.2$, $p=0.02$);
and robust location- and scale-invariant transformation ($\lambda = 1.1$, $\sum r_i = 59$, $\mu_{BC} = 2.4$, $\sigma_{BC} = 1.8$, $p=0.83$).

@@

The ischemic stroke dataset contains historic data from 126 patients with risk at ischemic stroke [@Kuhn2019-kt].
Applying conventional and invariant Box-Cox transformations to the maximum arterial wall thickness feature yielded the following results:
no transformation (residual sum $\sum r_i = 110$, $p=0.56$);
conventional transformation ($\lambda = -0.5$, $\sum r_i = 33$, $\mu_{BC} = 1.0$, $\sigma_{BC} = 0.2$, $p=0.01$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 1.1$, $\sum r_i = 127$, $\mu_{BC} = 5.5$, $\sigma_{BC} = 12$, $p=0.60$);
location- and scale-invariant transformation ($\lambda = -1.0$, $\sum r_i = 28$, $\mu_{BC} = 0.7$, $\sigma_{BC} = 0.1$, $p=0.01$);
and robust location- and scale-invariant transformation ($\lambda = 0.5$, $\sum r_i = 56$, $\mu_{BC} = 2.2$, $\sigma_{BC} = 1.4$, $p=0.35$).

@@

\begin{definition}[central portion of a sequence]
Let $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i \in \mathbb{R}\right\}$ be a finite sequence of length $n > 0$,
ordered so that $x_1 \leq x_2 \leq \ldots \leq x_n$.
Let $p_i = \frac{i - 1/3}{n + 1/3}$, with $i = 1, 2, \ldots n$ be the percentile value
corresponding to each element in $\mathbf{X}$. For elements with tied values in $\mathbf{X}$,
percentile values are replaced by the average in their group, i.e.
if $x_j = x_{j+1} = \ldots = x_{j+m}$, then 
$p_j' = p_{j+1}' = \ldots = p_{j+m}' = 1/(m+1)\sum_j^{j+m}p_j$.
Furthermore, let central portion $\kappa \in (0,1)$.

Then the central portion of $\mathbf{X}$ is 
$\mathbf{X}_{\text{c}} = \left\{x_i \in \mathbf{X} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}$.

\end{definition}

\begin{definition}[Residual errors of the central portion of a sequence]

Let the residual error for each element of $\mathbf{X}$ be $r_i =\left| \frac{x_i - \mu_M}{\sigma_M} - F^{-1}_{\mathcal{N}}(p_i) \right|$, 
with $\mu_M$ and $\sigma_M$ robust Huber M-estimates of location and scale of $\mathbf{X}$ [@Huber1981-su], 
and $F^{-1}_{\mathcal{N}}$ the quantile function of the normal distribution $\mathcal{N}(0, 1)$.

Then, the set of residual errors of the central portion of $\mathbf{X}$ (i.e. $\mathbf{X}_{\text{c}}$) is
$\mathbf{R}_{\text{c}} = \left\{ r_i \in \left\{ r_1, r_2, \ldots, r_n\right\} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}$.

\end{definition}

\begin{definition}[Central normality of a sequence]

The central portion of a sequence is normally distributed if the sum of residual
errors is equal to zero:
$\sum_{r_j \in \mathbf{R}_{\text{c}}} r_j = 0$

\end{definition}

In practice, a finite sequence sampled from a normal distribution $\mathcal{N}(\mu, \sigma)$ will have a non-zero sum of residual errors.

\begin{definition}[Central normality test]
The null-hypothesis is defined as: $\mathcal{H}_0: \sum_{r_j \in \mathbf{R}_{\text{c}}}r_j = 0$

THe alternative hypothesis is defined as: $\mathcal{H}_1: \sum_{r_j \in \mathbf{R}_{\text{c}}}r_j > 0$
\end{definition}

The null-hypothesis $\mathcal{H}_0$ is that the central portion of sequence $\mathbf{X}$
is normally distributed, with the alternative hypothesis $\mathcal{H}_1$ that it
is not normally distributed. To test this hypothesis, a test statistic is computed
and compared against a critical value.

\begin{definition}[Central normality test statistic]
Let $\alpha$ be the significance level for the central normality test.
Let $m$ be the number of elements of $\mathbf{R}_{\text{c}}$

The test statistic is then $\tau_{\alpha, n, \kappa} = \frac{1}{m}\sum_{r_j \in \mathbf{R}_{\text{c}}}r_j$.

\end{definition}

Critical values for the test statistic of the central normality test 
are found in Table \ref{tab:central-normality-critical-statistic}. These are 
determined by repeatedly and randomly sampling data (i.e. $\mathbf{X}$) from
a normal distribution $\mathcal{N}(0,1)$. The null-hypothesis should be rejected
at significance level $\alpha$ if 
$\tau_{\alpha, n, \kappa} \geq \tau_{\alpha, n, \kappa, \text{critical}}$.

The empirical central normality test is a central normality test, in the
definition provided above. However, critical values of the test statistic are
not determined directly from data drawn randomly from $\mathcal{N}(0,1)$.
Rather, after $\mathbf{X}$ is drawn, 10% of the values are randomly replaced by
random outlier values. The resulting distribution is centrally normal, but
not fully normal.

Critical values for the empirical variant of the central normality test are
found in Table \ref{tab:empirical-central-normality-critical-statistic}.

@@

@@

```{=latex}
\begin{table}
\begin{center}
\caption{Residual errors for features from real-world datasets after Yeo-Johnson transformation to normality. 
Several transformation methods include normalisation before tranformation, indicated by z-score normalisation (norm.) or robust scaling. 
AWT: arterial wall thickness; FE: fuel efficiency; PBM: penguin body mass}
\label{tab:normalisation-before-transformation-appendix}
\begin{tabular}{l | c c c c c}

@@

```{=latex}
\begin{table}
\begin{center}
\caption{Transformation parameter $\lambda$ for features from real-world datasets after Yeo-Johnson transformation to normality. 
Several transformation methods include normalisation before tranformation, indicated by z-score normalisation (norm.) or robust scaling. 
AWT: arterial wall thickness; FE: fuel efficiency; PBM: penguin body mass}
\label{tab:normalisation-before-transformation-lambda-appendix}
\begin{tabular}{l | c c c c c}

@@

```{=latex}
\begin{table}
\begin{center}
\caption{
p-values of empirical central normality tests for features from real-world datasets after Yeo-Johnson transformation to normality.
Several transformation methods include normalisation before tranformation, indicated by z-score normalisation (norm.) or robust scaling. 
AWT: arterial wall thickness; FE: fuel efficiency; PBM: penguin body mass}
\label{tab:normalisation-before-transformation-p-value-appendix}
\begin{tabular}{l | c c c c c}

@@

\midrule
none & 0.685 & 0.558 & 0.764 & 0.354 & 0.030 \\
conventional & 0.886 & 0.010 & 0.012 & 0.317 & 0.381 \\
conventional (z-score norm.) & 0.842 & 0.036 & 0.011 & 0.075 & 0.378 \\
conventional (robust scaling) & 0.855 & 0.002 & 0.007 & 0.077 & 0.416 \\
Raymaekers-Rousseeuw & 0.886 & 0.610 & 0.553 & 0.296 & 0.381 \\
Raymaekers-Rousseeuw (z-score norm.) & 0.685 & 0.632 & 0.812 & 0.075 & 0.378 \\
Raymaekers-Rousseeuw (robust scaling) & 0.826 & 0.609 & 0.813 & 0.077 & 0.434 \\
invariant & 0.907 & 0.134 & 0.026 & 0.049 & 0.549 \\
robust invariant & 0.781 & 0.102 & 0.583 & 0.085 & 0.717 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}
```

>>>>

data <- .plot_experimental_invariance(plot_theme = plot_theme, method = "box_cox")
data$plot
```

@@

Applying conventional and invariant Box-Cox transformations to age of patients with lung cancer [@Loprinzi1994-cd] yielded the following results:
no transformation (sum of residuals with normal distribution $\sum r_i = 16.5$, $p = 0.71$);
conventional transformation ($\lambda = 1.9$, $\sum r_i = 11.5$, $\mu_{BC} = 1.6 \cdot 10^3$, $\sigma_{BC} = 0.4 \cdot 10^3$, $p = 0.94$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 1.9$, $\sum r_i = 11.5$, $\mu_{BC} = 1.6 \cdot 10^3$, $\sigma_{BC} = 0.4 \cdot 10^3$, $p = 0.94$);
location- and scale-invariant transformation ($\lambda = 1.7$, $\sum r_i = 11.6$, $\mu_{BC} = 1.9$, $\sigma_{BC} = 0.8$, $p = 0.94$);
and robust location- and scale-invariant transformation ($\lambda = 1.5$, $\sum r_i = 11.6$, $\mu_{BC} = 3.6$, $\sigma_{BC} = 1.2$, $p = 0.93$).

@@

Applying conventional and invariant Box-Cox transformations to the body mass of penguins [@Gorman2014-eo] yielded the following results:
no transformation (residual sum $\sum r_i = 48.0$, $p < 0.001$);
conventional transformation ($\lambda = -0.5$, $\sum r_i = 32.2$, $\mu_{BC} = 2.1$, $\sigma_{BC} = 4 \cdot 10^{-3}$, $p = 0.06$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = -0.5$, $\sum r_i = 32.2$, $\mu_{BC} = 2.1$, $\sigma_{BC} = 4 \cdot 10^{-3}$, $p = 0.06$);
location- and scale-invariant transformation ($\lambda = 0.5$, $\sum r_i = 27.3$, $\mu_{BC} = 0.3$, $\sigma_{BC} = 0.6$, $p = 0.10$);
and robust location- and scale-invariant transformation ($\lambda = 0.2$, $\sum r_i = 25.2$, $\mu_{BC} = 0.4$, $\sigma_{BC} = 0.6$, $p = 0.44$).

@@

Applying conventional and invariant Box-Cox transformations to the latitude of houses in the Ames housing dataset [@De-Cock2011-jf] yielded the following results:
no transformation (residual sum $\sum r_i = 328$, $p < 0.001$);
conventional transformation ($\lambda = 62.1$, $\sum r_i = 319$, $\mu_{BC} = 1.1 \cdot 10^{99}$, $\sigma_{BC} = 0.0 \cdot 10^{99}$, $p < 0.001$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 96.0$, $\sum r_i = 319$, $\mu_{BC} = 6.2 \cdot 10^{153}$, $\sigma_{BC} = 0.3 \cdot 10^{153}$, $p < 0.001$);
location- and scale-invariant transformation ($\lambda = 1.9$, $\sum r_i = 312$, $\mu_{BC} = 2.3$, $\sigma_{BC} = 0.9$, $p < 0.001$);
and robust location- and scale-invariant transformation ($\lambda = 1.2$, $\sum r_i = 316$, $\mu_{BC} = 5.5$, $\sigma_{BC} = 1.4$, $p < 0.001$).

@@

data <- .plot_experimental_outlier_robustness(plot_theme = plot_theme, method = "box_cox")
data$plot
```

@@

The Top Gear dataset contains data on 297 vehicles, with outliers related highly fuel-efficient vehicles [@Alfons2021-kc].
Applying conventional and invariant Box-Cox transformations to the fuel consumption feature yielded the following results:
no transformation (residual sum $\sum r_i = 54$, $p=0.58$);
conventional transformation ($\lambda = -0.1$, $\sum r_i = 55$, $\mu_{BC} = 3.0$, $\sigma_{BC} = 0.3$, $p < 0.001$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 0.8$, $\sum r_i = 48$, $\mu_{BC} = 29$, $\sigma_{BC} = 15$, $p=0.32$);
location- and scale-invariant transformation ($\lambda = -0.7$, $\sum r_i = 44$, $\mu_{BC} = 0.6$, $\sigma_{BC} = 0.2$, $p < 0.001$);
and robust location- and scale-invariant transformation ($\lambda = 1.1$, $\sum r_i = 59$, $\mu_{BC} = 2.4$, $\sigma_{BC} = 1.8$, $p=0.67$).

@@

The ischemic stroke dataset contains historic data from 126 patients with risk at ischemic stroke [@Kuhn2019-kt].
Applying conventional and invariant Box-Cox transformations to the maximum arterial wall thickness feature yielded the following results:
no transformation (residual sum $\sum r_i = 110$, $p=0.86$);
conventional transformation ($\lambda = -0.5$, $\sum r_i = 33$, $\mu_{BC} = 1.0$, $\sigma_{BC} = 0.2$, $p=0.01$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 1.1$, $\sum r_i = 127$, $\mu_{BC} = 5.5$, $\sigma_{BC} = 12$, $p=0.90$);
location- and scale-invariant transformation ($\lambda = -1.0$, $\sum r_i = 28$, $\mu_{BC} = 0.7$, $\sigma_{BC} = 0.1$, $p=0.02$);
and robust location- and scale-invariant transformation ($\lambda = 0.5$, $\sum r_i = 56$, $\mu_{BC} = 2.2$, $\sigma_{BC} = 1.4$, $p=0.60$).

@@

Critical values for the central normality test and its empirical variant are
found in Table \ref{tab:central-normality-critical-statistic} and Table 
\ref{tab:empirical-central-normality-critical-statistic}, respectively.

@@

```{r empirical-central-normality-critical-statistics, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=2.5}
cap <- paste0(
  "Critical test statistic $\\tau_{\\alpha, n, \\kappa = 0.70}$ of the (empirical) ",
  "central normality test as function of $n$ for several values of significance level $\\alpha$. ",
  "The critical test statistics for central normality test are determined using fully normal data, ",
  "whereas the statistics for the empirical variant are determined using centrally normal data, ",
  "i.e. with fully normal data where 10% of elements are replaced by outliers. ",
  "emp: empirical"
)

.plot_test_statistic_tau(
  plot_theme = plot_theme,
  manuscript_dir = manuscript_dir
)

@@

```{=latex}
\begin{table}
\begin{center}
\caption{Residual errors for features from real-world datasets after Yeo-Johnson transformation to normality. 
Several transformation methods include normalisation before tranformation, indicated by z-score normalisation (norm.) or robust scaling. 
AWT: arterial wall thickness; FE: fuel efficiency; PBM: penguin body mass}
\label{tab:normalisation-before-transformation-appendix}
\begin{tabular}{l | r r r r r}

@@

```{=latex}
\begin{table}
\begin{center}
\caption{Transformation parameter $\lambda$ for features from real-world datasets after Yeo-Johnson transformation to normality. 
Several transformation methods include normalisation before tranformation, indicated by z-score normalisation (norm.) or robust scaling. 
AWT: arterial wall thickness; FE: fuel efficiency; PBM: penguin body mass}
\label{tab:normalisation-before-transformation-lambda-appendix}
\begin{tabular}{l | r r r r r}

@@

```{=latex}
\begin{table}
\begin{center}
\caption{
p-values of empirical central normality tests for features from real-world datasets after Yeo-Johnson transformation to normality.
Several transformation methods include normalisation before tranformation, indicated by z-score normalisation (norm.) or robust scaling. 
AWT: arterial wall thickness; FE: fuel efficiency; PBM: penguin body mass}
\label{tab:normalisation-before-transformation-p-value-appendix}
\begin{tabular}{l | r r r r r}

@@

\midrule
none                                  & 0.712 & 0.864 & 0.577     & $< 0.001$ & $< 0.001$ \\
conventional                          & 0.936 & 0.017 & $< 0.001$ & $< 0.001$ & 0.058 \\
conventional (z-score norm.)          & 0.860 & 0.058 & $< 0.001$ & $< 0.001$ & 0.054 \\
conventional (robust scaling)         & 0.881 & 0.001 & $< 0.001$ & $< 0.001$ & 0.076 \\
Raymaekers-Rousseeuw                  & 0.936 & 0.905 & 0.320     & $< 0.001$ & 0.058 \\
Raymaekers-Rousseeuw (z-score norm.)  & 0.605 & 0.922 & 0.635     & $< 0.001$ & 0.054 \\
Raymaekers-Rousseeuw (robust scaling) & 0.830 & 0.948 & 0.637     & $< 0.001$ & 0.085 \\
invariant                             & 0.948 & 0.231 & $< 0.001$ & $< 0.001$ & 0.184 \\
robust invariant                      & 0.797 & 0.182 & 0.352     & $< 0.001$ & 0.397 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}
```
